---
title: "Mapping bike crashes with GIS-tools using the R-command line"
author: "Sighvatur S. Davidsson"
date: '2022-02-10'
---

```{r document, echo= FALSE}
# Supress warnings
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 

# Makes sure error messages get printed in english
Sys.setenv(lang = "en_US")

options(scipen = 999,
        blogdown.knit.on_save = FALSE # Mapping can take some time. Make sure we're not knitting '.Rmd' to '.html' everytime we click save
        )

# blogdown::check_site() # Note to self: Try running this if anything goes wrong.

```

# This is part 2 of 4 total in a series of working with bike crash data. In the previous post I introduced the *brike crash*-dataset and concluded that it has a very qualitative nature. I also did some quick data visualisation of the crash locations. 

1. [Introducing the bike crash-dataset](https://sdavidsson.netlify.app/post/2022-01-24-introducing-the-bike-crash-data/)
2. Mapping bike crashes with GIS-tools using the R-command line
3. Identifying environmental factors that affect the number of crashes
4. How can these insights make an impact? 

In this post we will be focusing on mapping the cases with GIS (Geographic Information Systems) tools using the R command line. Having worked with GIS-technology in ArcGIS and Q-GIS before, I am fascinated by how much is possible in R. In my opinion the SF-package is the best thing that has happened for R since dplyr. It relies on a PostGIS standard to enable cross platform work and feels very much in line with the tidyverse-dialect we know from the essential modern R functions. 


```{r Workspace, echo = TRUE}
# Load packages
pacman::p_load(tidyverse, sf, tmap, plotly)

# Read in datasets
bike_crash <- read_rds("~/R/Projects/bike_crash/processed_data/bike_crash.rds")

# Read in North Carolina counties geometry
nc_counties <- read_rds("~/R/Projects/bike_crash/processed_data/nc_counties_geom.rds")
```

##### What are we mapping? 

```{r}
bike_crash %>% 
  count(crash_sevr)




  ggplotly() 



```

At first we will be mapping all crashes that have taken place in the period 2007 - 2019


We are interested in crash locations
- all crashes
- in the period 2007 - 2019


##### First map of crash locations

```{r}
tmap_mode("plot") 

# Etablering af geometrier
bike_crash <- st_as_sf(bike_crash, coords = c(x = "longitude", y = "latitude"), crs = "NAD83") 

# Initial map of crash locations
bike_crash %>%
  tm_shape() + 
  tm_dots(alpha = 0.1) + 
  tm_shape(nc_counties) + 
  tm_borders(alpha = 0.4, lwd = 1.3) + 
  tm_layout(frame = FALSE,
            main.title = "Crashes in North Carolina 2007-2019", 
            main.title.position = "center",
            main.title.fontfamily = "IBM Plex Sans",
            main.title.size = 1,
            ) 
```


### Hetmapping with the Two-dimensional Kernel Density Estimator (2D KDE)

Computer graphics rely on two different technologies - vectors and raster data. The previous map 

raster data (can be thought of as a pixel mosaic)

The 2D-KDE utilises the latter, and thus 


```{r}
# We specify the number of quadrats (the level of detail in our raster map)
n_quadrats <- 100

# Bestemmelse af optimal bandwidth (gennemsnittet af to én-dimensionelle)
bandwidth <- mean(c(MASS::bandwidth.nrd(bike_crash$x), MASS::bandwidth.nrd(bike_crash$y)))

# Her køres metoden 
kde <- MASS::kde2d(bike_crash$x, 
                   bike_crash$y, 
                   h = bandwidth, 
                   n = n_quadrats)

# Og vi konverterer til raster (for at kunne mappe)
kde <- raster::raster(kde)

# _ _ _
# MAPPING
tmap_mode("plot") 

# Vi vil gerne have heatmaps på en hvid baggrund (det er pænere)
col_palette <- c("#FFFFFF", "#fee391", "#fec44f", "#fe9929", "#ec7014", "#cc4c02", "#8c2d04")

# Vi vil også gerne kunne fremhæve der, hvor de fleste sammenstød sker
top_counties <- nc_counties %>% filter(county %in% c("Mecklenburg County", "Wake County", "Durham County", "New Hanover County", "Guilford County"))


tm_shape(nc_counties) + 
  tm_polygons(alpha = 0) + 
tm_shape(kde) + 
  tm_raster("layer",  
            palette = col_palette, 
            alpha = .7) + 
tm_shape(nc_counties) + 
  tm_polygons(alpha = 0, 
              lwd = 1.3,
              id = "county") +
  tm_layout(main.title = "Områder med den største tæthed af sammenstød", 
            main.title.position = "center",
            main.title.size = .9,
            frame = TRUE, 
            legend.show = FALSE
            ) +
  tm_shape(top_counties) +
  tm_text("county",
          just = "left",
          scale = 0.75,
          shadow = T,
          clustering = F) + 
  tm_credits("Bandwidth = 0.68 (metodespecificeret)", position = c("right", "bottom"), align = "right")
```


